Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-05-19 introduction.tex
Added most of Dave Collins' suggested edits (except for a bunch of stuff having to do with the hydro).

Therefore, we would like to make use of the expertise that has been invested in Eulerian solvers; unfortunately, as originally developed, none of these schemes provide an easy way to adaptively and dynamically increase the spatial and temporal resolution in small volumes, as required by, for example, gravitational instability.  We have instead adopted an idea first suggested by \citet{Berger89} in the Computational Fluid Dynamics (CFD) community, which has become known as Structured Adaptive Mesh Refinement (SAMR).  The principle is to adaptively add and modify additional, finer meshes (``patches'') over regions that require higher resolution.  Since each patch has a simple geometry (Cartesian, in our case) the hyperbolic fluid-dynamical equations can be solved on that patch with a well-developed `off-the-shelf' numerical method.  In addition, it is relatively straightforward to add other advanced physics (e.g. magnetohydrodynamics, radiation transport, chemistry, etc.). This paper is devoted to describing how we have implemented this idea for astrophysics, including the addition of comoving coordinates, self-gravity, radiative cooling, chemistry, heat conduction, collisionless fluids, MHD, radiation transport, star formation and a range of other physical effects.

@@

\enzo\ is a structured adaptive mesh refinement (SAMR) code that was originally developed for cosmological hydrodynamics and has been used on a wide variety of problems.  It has been expanded as a general tool for astrophysical fluid dynamics and is intended to be efficient, accurate and easily expanded.  Although many of the components of the code have been described in previous publications \citep{1995CoPhC..89..149B, BryanThesis96, Bryan97a, Bryan97b, Norman99, BryanCompSci99, Bryan01, Oshea04, 2007arXiv0705.1556N}, there has been no systematic and complete description of the code.  In this paper we provide that description, filling in many gaps and showing the code's performance for a wide variety of test problems.

The \enzo\ code has been extensively used over the last two decades in a wide variety of problems, and more than 100 peer-reviewed papers have been published based on results obtained with the code.  The code has previously been used to study a wide range of astrophysical systems, including galaxy clusters \citep{Loken02}, the interstellar medium \citep{Slyz05}, the intergalactic medium \citep{Fang01}, cooling flows \citep{Li12}, turbulence \citep{Kritsuk04}, and the formation of the first stars \citep{ABN02}.

>>>>

Therefore, we would like to make use of the expertise that has been invested in Eulerian solvers; unfortunately, as originally developed, none of these schemes provide an easy way to adaptively and dynamically increase the spatial and temporal resolution in small volumes, as required by, for example, gravitational instability.  We have instead adopted an idea first suggested by \citet{Berger89} in the Computational Fluid Dynamics (CFD) community, which has become known as Structured Adaptive Mesh Refinement (SAMR).  The principle is to adaptively add and modify additional, finer meshes (``grids'') over regions that require higher resolution.  Since each patch has a simple geometry (Cartesian, in our case) the hyperbolic fluid-dynamical equations can be solved on that patch with a well-developed `off-the-shelf' numerical method.  In addition, it is possible to add other advanced physics (e.g. magnetohydrodynamics, radiation transport, chemistry, etc.). This paper is devoted to describing how we have implemented this idea for astrophysics, including the addition of comoving coordinates, self-gravity, radiative cooling, chemistry, heat conduction, collisionless fluids, MHD, radiation transport, star formation and a range of other physical effects.

@@

\enzo\ is a structured adaptive mesh refinement (SAMR) code that was originally developed for cosmological hydrodynamics and has been used on a wide variety of problems.  It has been expanded as a general tool for astrophysical fluid dynamics and is intended to be efficient, accurate and easily expanded.  Although many of the components of the code have been described in previous publications \citep{1995CoPhC..89..149B, BryanThesis96, Bryan97a, Bryan97b, Norman99, BryanCompSci99, Bryan01, Oshea04, 2007arXiv0705.1556N, WangAbelZhang08,ReynoldsHayesPaschosNorman2009,Collins10,Wise11_Moray}, there has been no systematic and complete description of the code.  In this paper we provide that description, filling in many gaps and showing the code's performance for a wide variety of test problems.

The \enzo\ code has been extensively used over the last two decades in a wide variety of problems, and more than 100 peer-reviewed papers have been published based on results obtained with the code.  The code has previously been used to study a wide range of astrophysical systems, including galaxies \citep{2003ApJ...587...13T, 2012ApJ...749..140H}, galaxy clusters \citep{Loken02, Skillman13}, the interstellar medium \citep{Slyz05}, the intergalactic medium \citep{Fang01, 2011ApJ...731....6S}, cooling flows \citep{Li12, 2013ApJ...763...38S}, turbulence \citep{Kritsuk04, 2009JPhCS.180a2020K, Collins11}, the formation of the first stars \citep{ABN02, 2007ApJ...654...66O, 2009Sci...325..601T}, and the formation of stars in our own Galaxy \citep{Collins12a}.

### 2013-05-19 ms.tex
Rearranging sections - cleaned this up a bit.
Added most of Dave Collins' suggested edits (except for a bunch of stuff having to do with the hydro).

% Paper Outline  (last updated January 18, 2013)
%
% 1.  Introduction
% 2.  Physical equations and overview
%   2.1 Physical equations
%   2.2 Overview of numerical methods
% 3.  Numerical methods   %%%%%% THIS IS NOW A BUNCH OF DIFFERENT SECTIONS
%   3.1   AMR
%   3.2.  Hydro: Godunov (PPM)
%   3.3   Hydro: Godunov (MUSCL)
%   3.4   Hydro: Finite difference (ZEUS)
%   3.5   MHD: Dedner
%   3.6   MHD: CT
%   3.7   Gravity
%   3.8   Particle Integration
%   3.9.  Chemistry
%   3.10 Cooling
%   3.11 Radiation: Homogeneous
%   3.12 Radiation: ray-tracing
%   3.13 Radiation: FLD
%   3.14 Thermal conduction
%   3.15 Star formation & Feedback
%   3.16 Timestep constraints
%   3.17 Analysis
% 4. Code tests
%   4.1 Verifying and Validating the Enzo code
%   4.2 Representative test problems
% 5. Parallelism 
%   5.1 Parallel strategy 
%   5.2 Parallel performance 
% 6. Code development methodology
% 7. Conclusions
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{abstract}
This paper describes the open-source \enzo\ code, which uses a
block-structured adaptive mesh refinement to provide high spatial and
temporal resolution for modeling astrophysical fluid flows.  The code
is Cartesian and can be run in 1, 2, and 3 dimensions, and supports a
large variety of physics, including hydrodynamics, ideal and non-ideal
magnetohydrodynamics, N-body dynamics (and, more broadly, self-gravity
of fluids and particles), primordial gas chemistry, optically-thin
radiative cooling of primordial and metal-enriched plasmas, radiation
transport, and cosmological expansion.  In addition to explaining the
algorithms implemented, we present solutions for a wide range of test
problems, demonstrate the code's parallel scaling properties, and
discuss the \enzo\ consortium's code development methodology.
\end{abstract}

@@

\section{Fluid methods}
\label{sec:fluids}

@@

\section{Gravity and N-body}
\label{sec:allgrav}
\input{numerical-gravity}
\input{numerical-particle}

\section{Microphysics}
\label{sec:microphysics}
\input{numerical-chemistry}
\input{numerical-cooling}

\section{Radiation}
\label{sec:allrad}
\input{numerical-radiation-homogeneous} 
\input{numerical-radiation-raytracing} 
\input{numerical-radiation-fld}

\section{Other physics}
\label{sec:otherphys}
\input{numerical-conduction} 
\input{numerical-starformation}

>>>>

% Paper Outline  (last updated May 17, 2013)
%
% 1.  Introduction
% 2.  Physical equations and overview of numerical methodology
%   2.1 Physical equations
%   2.2 Overview of numerical methods
% 3.  The structured AMR method
% 4.  Fluid methods
%   4.1 PPM
%   4.2 MUSCL
%   4.3 MHD-CT
%   4.4 ZEUS 
% 5.  Gravity and N-body
%   5.1 Gravity
%   5.2 N-body dynamics
% 6. Microphysics
%   6.1  Chemistry
%   6.2  Radiative Cooling and heating
% 7. Radiation
%   7.1 Radiation: Homogeneous
%   7.2 Radiation: ray-tracing
%   7.3 Radiation: FLD
% 8. Other physics
%   8.1 Thermal conduction
%   8.2 Star formation & Feedback
% 9. Time stepping
% 10.  Analysis
% 11. Code tests
%   11.1 Verifying and Validating the Enzo code
%   11.2 Representative test problems
% 12. Parallelism 
%   12.1 Parallel strategy 
%   12.2 Parallel performance 
% 13. Code development methodology
% 14. Conclusions
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{abstract}
This paper describes the open-source code \enzo, which uses a
block-structured adaptive mesh refinement to provide high spatial and
temporal resolution for modeling astrophysical fluid flows.  The code
is Cartesian and can be run in 1, 2, and 3 dimensions, and supports a
large variety of physics, including hydrodynamics, ideal and non-ideal
magnetohydrodynamics, N-body dynamics (and, more broadly, self-gravity
of fluids and particles), primordial gas chemistry, optically-thin
radiative cooling of primordial and metal-enriched plasmas, radiation
transport, and cosmological expansion.  In addition to explaining the
algorithms implemented, we present solutions for a wide range of test
problems, demonstrate the code's parallel scaling properties, and
discuss the \enzo\ collaboration's code development methodology.
\end{abstract}

@@

\section{Fluid methods}
\label{sec.fluids}

@@

\section{Gravity and N-body}
\label{sec.allgrav}
\input{numerical-gravity}
\input{numerical-particle}

\section{Microphysics}
\label{sec.microphysics}
\input{numerical-chemistry}
\input{numerical-cooling}

\section{Radiation}
\label{sec.allrad}
\input{numerical-radiation-homogeneous} 
\input{numerical-radiation-raytracing} 
\input{numerical-radiation-fld}

\section{Other physics}
\label{sec.otherphys}
\input{numerical-conduction} 
\input{numerical-starformation}

### 2013-05-19 numerical-amr.tex
Rearranging sections.
Added most of Dave Collins' suggested edits (except for a bunch of stuff having to do with the hydro).

\subsection{The Structured Adaptive Mesh Refinement (SAMR) Method}
\label{sec.amr}

@@

Each grid patch in \enzo\ contains arrays of values for baryon and particle quantities.   Grids are partitioned into a core of \emph{real zones} and a surrounding layer of \emph{ghost zones}, as shown in Figure~\ref{fig:amr_hierarchy}.  The real zones store field values and ghost zones are used to temporarily store values that have been obtained directly from neighboring grids or interpolated from a parent grid.  These zones are necessary to accommodate the computational stencil of the (magneto)hydrodynamics solvers (Sections~\ref{sec.hydro.ppm} and~\ref{sec.hydro.zeus}) and the gravity solver (Section~\ref{sec.gravity}).  The PPM and Zeus hydro solvers require 3 layers of ghost zones,  and the gravity solver requires 6 (although only for the density and potential fields).  This can lead to significant memory and computational overhead, particularly for smaller grid patches at high levels of refinement.

From the point of view of the hydro solver, each grid is solved as an independent CFD problem, with Dirichlet boundary conditions stored in the ghost zones.  At the end of each time step, cells on a given level fill ghost zones by interpolating from the parent and copying from neighbors on the same level.  Grids then correct the flux at the interface boundary between fine and coarse zones, and finally project their real zone data to the parent grid.

@@

\subsubsection{Setting the boundary values}
\label{sec:interpolation}

@@

In the first case, we allow for four types of boundary conditions:
\begin{enumerate}
  \item{\em Reflecting.} In this case, the boundary behaves like a mirror, with the solution on the boundary reflecting the solution in the computational domain ($q(-x) = q(+x)$) but with the velocity normal to the boundary reversed: $v_x(-x) = -v_x(+x)$.
  \item{\em Outflow.} True outflow conditions are extremely difficult \red{(REF?)}.  We approximate them by duplicating the solution at the edge of the computational domain: $q(-x) = q(0)$.
  \item{\em Inflow.} In this case, the boundary values are fixed by a pre-determined function ($q(-x) = q0(x,t)$).  The code provides a simple way to set inflow values that are constant in both time and space over an entire face, or hooks are provided for special cases where the boundary must be time-varying.
  \item{\em Periodic.} The boundary solution is obtained from the other side of the grid: $q(-x) = q(x_{\rm{max}}-x)$.
\end{enumerate}

@@

\subsubsection{Solving the hydrodynamic equations on a grid}
\label{sec:solve_hydro}

Given the grid and its boundary conditions, the process of solving
equations~(\ref{eq:mass})--(\ref{eq:total_energy}) can be detached
from the rest of the SAMR technique, so that it is possible to
experiment with a number of different methods.  The restrictions are
that the method should be flux-conservative, or at least produce an
accurate estimate of the flux of conserved quantities at cell
interfaces.  This is required in the flux-correction step (see
\ref{sec:flux_correction}).  There is also the issue of centering,
which refers to the position of physical quantities within a cell (the
primary types are cell-centered, face-centered, and node-centered). We
will employ both the direct Eulerian version of the piecewise
parabolic method \citep{1984JCoPh..54..174C,1995CoPhC..89..149B} which
assumes cell-centering for all quantities (actually they are
cell-averages).  However, it is also possible to use other schemes,
which employ different centering for different variables
\citep[e.g.,][]{Stone92a}.  The interpolation step, described in
Section~\ref{sec:interpolation}, must correctly interpolate with the
selected centering.

\subsubsection{Projection}
\label{sec:projection}

Since a refined region is simulated with (at least) two different resolutions (one fine and one coarse), it has two or
more ``solutions'' (i.e. values of density, velocity, etc.) that represent the same volume of space.  This is one important aspect of the SAMR method -- some physical regions are represented by grids at multiple resolutions.  To
maintain consistency, the coarse grid quantities are updated with the finer values, using a volume-weighted average of the conserved quantities:
\begin{equation}
q^{\rm coarse} = r^{-d} \sum q^{\rm fine}_{i, j,k}
\end{equation}
where $r$ is the refinement factor, $d$ the dimensionality, and the sum is over all fine cells that ``cover" the coarse cell.

\subsubsection{Flux correction}
\label{sec:flux_correction}

One of the significant advantages of flux-conservative methods is, of course, their ability to maintain conserved quantities to machine precision.  Unfortunately, the projection step upsets this flux conservation since, at the boundary of a refined region, the coarse and fine cells on either side of the border are updated with different estimates of the flux across the boundary.  In one case this comes from the coarse grid computation, and in the other from the fine grid solution.  Conservation can be restored by correcting all coarse cells that lie at the boundary of a fine region
with the difference between the coarse and fine estimates of the flux across the boundary:

@@

\subsubsection{Hierarchy reconstruction}
\label{sec:hierarchy_reconstruction}

Since the addition of more highly refined grids is adaptive, the conditions for refinement must be specified.  Up to seven different criteria can be specified for a given simulation, and these refinement criteria are discussed in more detail in the following section (Section~\ref{sec:refinement_criteria}).   Once all cells in a given grid that require additional refinement have been flagged, rectangular boundaries are determined that minimally encompass the flagged cells.  This is done using an iterative machine-vesion technique as described in \red{(REF)}.

We first find the smallest region (or ``proto-subgrid") that covers all of the refined cells on a grid.  We compute an efficiency $\epsilon$ for this grid, defined as the ratio of flagged cells to total cells.  If this efficiency is sufficiently larger (typically larger than 30\%), or if the grid is small enough (typically less $4^d$ coarse cells) then we adopt the region as a new grid and move on to the next region (if any).  If the region is not acceptable, then we try to split it into two using the following algorithm.  First, we determine the longest dimension and compute the signature along that dimension.  The signature is found by summing the number of refined cells with the same index along the longest dimension.  Edges are typically indicates by inflection points in this signature, so we calculate the second derivative using a simple three-point finite difference estimator and look for the maximum in the second derivative, with ties going to the index closest to the center.  The original region is then split into two new regions, or proto-subgrids.  This process is then re-applied to each of the new regions.

This algorithm is not necessarily optimal, but experience shows it does a reasonable job of generating an efficient covering of the refined cells.  Note that some cells are unnecessarily refined with the SAMR approach, resulting in wasted resources (unlike in cell-splitting AMR schemes).  However, it has a number of advantages.  First, traditional grid-based hydro method can be simply added with only minor modifications.  Second, it is easier to accommodate larger stencils (as in PPM, which requires three boundary zones), and finally it can be more efficient for some machines since there are more predictable and linear memory accesses, which more efficiently use a CPU's cache.

\subsubsection{Refinement criteria}
\label{sec:refinement_criteria}

Finally, we turn to the refinement criteria themselves.  Although generalized refinement criteria are available based on estimators for the error, such as the Richardson method \red{(REF)}, we have focused generally on methods that use physical estimators based on the assumption that the simulator has a specific quantity or quantities that indicate the need for refinement.  The flagging methods currently built into \enzo\ are:

@@

\item{\em Shocks.}  In this method, we identify and refine cells that contain shocks using the following criteria:
\begin{eqnarray}
(p_{i+1} - p_{i-1})/\min(p_{i+1}, p_{i-1}) & > & 0.33  \nonumber \\
u_{i-1} - u_{i+1} & > & 0 \\
e_i / E_i & > & 0.1  \nonumber
\end{eqnarray}

\item{\em Particle mass.}  This refinement criterion is precisely the same as the baryon overdensity criterion except that it uses the gridded (CIC) particle density.  In general, this criteria is used in cosmological simulations, and thus refines on dark matter and stellar density.

@@

where $\sigma_{\rm HI}$ is the neutral atomic hydrogen absorption
cross section, $n_{\rm HI}$ is the proper \ion{H}{1} number density,
and $\Delta x$ is the cell width.  If $\tau > 1$, the cell is refined
further, up to the maximum level of refinement.

@@

\item{\em ``Must refine region.''}  This refinement criterion is used to ensure that all cells in a given subvolume of the simulation are refined to {\it at least} some minimum level of refinement.  The subvolume is user-specified, and may evolve over time (e.g., to follow a structure of interest that is traveling through the simulation volume).

>>>>

\section{The Structured Adaptive Mesh Refinement (SAMR) Method}
\label{sec.amr}

@@

Each grid patch in \enzo\ contains arrays of values for baryon and particle quantities.   Grids are partitioned into a core of \emph{active zones} and a surrounding layer of \emph{ghost zones}, as shown in Figure~\ref{fig:amr_hierarchy}.  The active zones store field values and ghost zones are used to temporarily store values that have been obtained directly from neighboring grids or interpolated from a parent grid.  These zones are necessary to accommodate the computational stencil of the (magneto)hydrodynamics solvers (Sections~\ref{sec.hydro.ppm} and~\ref{sec.hydro.zeus}) and the gravity solver (Section~\ref{sec.gravity}).  The PPM and Zeus hydro solvers require 3 layers of ghost zones,  and the gravity solver requires 6 (although only for the density and potential fields).  This can lead to significant memory and computational overhead, particularly for smaller grid patches at high levels of refinement.

From the point of view of the hydro solver, each grid is solved as an independent CFD problem, with Dirichlet boundary conditions stored in the ghost zones.  At the end of each time step, cells on a given level fill ghost zones by interpolating from the parent and copying from neighbors on the same level.  Grids then correct the flux at the interface boundary between fine and coarse zones, and finally project their active zone data to the parent grid.

@@

\subsection{Setting the boundary values}
\label{sec:interpolation}

@@

In the first case, we allow for four types of boundary conditions:
\begin{enumerate}
  \item{\em Reflecting.} In this case, the boundary behaves like a mirror, with the solution on the boundary reflecting the solution in the computational domain ($q(-x) = q(+x)$) but with the velocity normal to the boundary reversed: $v_x(-x) = -v_x(+x)$.
  \item{\em Outflow.}  We approximate outflow boundary conditions by duplicating the solution at the edge of the computational domain: $q(-x) = q(0)$.
  \item{\em Inflow.} In this case, the boundary values are fixed by a pre-determined function ($q(-x) = q0(x,t)$).  The code provides a simple way to set inflow values that are constant in both time and space over an entire face, or hooks are provided for special cases where the boundary must be time-varying.
  \item{\em Periodic.} The boundary solution is obtained from the other side of the grid: $q(-x) = q(x_{\rm{max}}-x)$.
\end{enumerate}

@@

\subsection{Solving the hydrodynamic equations on a grid}
\label{sec:solve_hydro}

Given the grid and its boundary conditions, the process of solving
equations~(\ref{eq:mass})--(\ref{eq:total_energy}) can be detached
from the rest of the SAMR technique, so that it is possible to
experiment with a number of different methods.  The restrictions are
that the method should be flux-conservative, or at least produce an
accurate estimate of the flux of conserved quantities at cell
interfaces.  This is required in the flux-correction step (see
\ref{sec:flux_correction}).  There is also the issue of centering,
which refers to the position of physical quantities within a cell (the
primary types are cell-centered, face-centered, and node-centered). We
will employ the direct Eulerian version of the piecewise
parabolic method \citep{1984JCoPh..54..174C,1995CoPhC..89..149B}, which
assumes cell-centering for all quantities (actually they are
cell-averages).  However, it is also possible to use other schemes,
which employ different centering for different variables
\citep[e.g.,][]{Stone92a}.  The interpolation step, described in
Section~\ref{sec:interpolation}, must correctly interpolate with the
selected centering.

\subsection{Projection}
\label{sec:projection}

Since a refined region is simulated with (at least) two different resolutions (one fine and one coarse), it has two or
more ``solutions'' (i.e. values of density, velocity, etc.) that represent the same volume of space.  This is one important aspect of the SAMR method -- some physical regions are represented by grids at multiple resolutions.  To
maintain consistency, the coarse grid quantities are updated with the finer values, using a volume-weighted average of the conserved quantities:
\begin{equation}
q^{\rm coarse} = r^{-d} \sum q^{\rm fine}_{i, j,k}
\end{equation}
where $r$ is the refinement factor and the sum is over all fine cells that ``cover" the coarse cell. The exponent, $d$, refers to the weight of the average.  For cell-centered quantities, this is the dimensionality of the problem.  For face-centered quantities, this is one less than the dimensionality, and for edge-centered quantities this is one.

\subsection{Flux correction}
\label{sec:flux_correction}

One of the significant advantages of flux-conservative methods is, of course, their ability to maintain conserved quantities to machine precision.  Unfortunately, the projection step upsets this flux conservation since, at the boundary of a refined region, the coarse and fine cells on either side of the border are updated with different estimates of the flux across the boundary.  At one refinement level this comes from the coarse grid computation, and in the other from the fine grid solution.  Conservation can be restored by correcting all coarse cells that lie outside the boundary of a fine region
with the difference between the coarse and fine estimates of the flux across the boundary:

@@

\subsection{Hierarchy reconstruction}
\label{sec:hierarchy_reconstruction}

Since the addition of more highly refined grids is adaptive, the conditions for refinement must be specified. Many different criteria can be simultaneously specified for a given simulation, and these refinement criteria are discussed in more detail in the following section (Section~\ref{sec:refinement_criteria}).   Once all cells in a given grid that require additional refinement have been flagged, rectangular boundaries are determined that minimally encompass the flagged cells.  This is done using an iterative machine-vesion technique as described in \red{(REF)}.

We first find the smallest region (or ``proto-subgrid") that covers all of the refined cells on a grid.  We compute an efficiency $\epsilon$ for this grid, defined as the ratio of flagged cells to total cells.  If this efficiency is sufficiently larger (typically larger than 30\%), or if the grid is small enough (typically less than $4^d$ coarse cells, where d is the dimensionality of the simulation) then we adopt the region as a new grid and move on to the next region (if any).  If the region is not acceptable, then we split it into two using the following algorithm.  First, we determine the longest dimension and compute the signature along that dimension.  The signature is found by summing the number of refined cells with the same index along the longest dimension.  Edges are typically indicates by inflection points in this signature, so we calculate the second derivative using a simple three-point finite difference estimator and look for the maximum in the second derivative, with ties going to the index closest to the center.  If no inflection points are found along the longest exis, the next-longest axis is searched.  If no inflection points are found, the longest axis is cut in half.  The original region is then split into two new regions, or proto-subgrids.  This process is then re-applied to each of the new regions.

This algorithm is not necessarily optimal, but experience shows it does a reasonable job of generating an efficient covering of the refined cells.  Note that some cells are unnecessarily refined with the SAMR approach, resulting in wasted resources (unlike in cell-splitting AMR schemes).  However, it has a number of advantages.  First, traditional grid-based hydro method can be simply added with only minor modifications.  Second, it is easier to accommodate larger stencils (as in PPM, which requires three boundary zones), and third, it can be more efficient for some machines since there are more predictable and linear memory accesses, which more efficiently use a CPU's cache.

\subsection{Refinement criteria}
\label{sec:refinement_criteria}

In this section, we turn to the refinement criteria themselves.  Although generalized refinement criteria are available based on estimators for the error, such as the Richardson method \red{(REF)}, we have focused generally on methods that use physical estimators based on the assumption that the simulator has a specific quantity or quantities that indicate the need for refinement.  The flagging methods currently built into \enzo\ are:

@@

\item{\em Shocks.}  In this method, we identify and refine cells that contain shocks using the following criteria:
\begin{eqnarray}
(p_{i+1} - p_{i-1})/\min(p_{i+1}, p_{i-1}) & > & 0.33  \nonumber \\
u_{i-1} - u_{i+1} & > & 0 \\
e_i / E_i & > & 0.1  \nonumber
\end{eqnarray}
We note that the values $0.33$ and $0.1$ in the equations above were determined empirically, and can be modified by the user to refine more or less aggressively, as the situation warrants.

\item{\em Particle mass.}  This refinement criterion is precisely the same as the baryon overdensity criterion except that it uses the gridded (cloud-in-cell, or CIC) particle density.  In general, this criteria is used in cosmological simulations, and thus refines on dark matter and stellar density.

@@

where $\sigma_{\rm HI}$ is the neutral atomic hydrogen absorption
cross section, $n_{\rm HI}$ is the proper \ion{H}{1} number density,
and $\Delta x$ is the cell width (which is assumed to be the same along each axis).  If $\tau > 1$, the cell is refined
further, up to the maximum level of refinement.

@@

\item{\em Must refine region.}  This refinement criterion is used to ensure that all cells in a given subvolume of the simulation are refined to {\it at least} some minimum level of refinement.  The subvolume is user-specified, and may evolve over time (e.g., to follow a structure of interest that is traveling through the simulation volume).

### 2013-05-19 numerical-analysis.tex
Rearranging sections.
Added most of Dave Collins' suggested edits (except for a bunch of stuff having to do with the hydro).

\subsection{Analysis}
\label{sec.num.analysis}

\subsubsection{Inline analysis with yt}

Analyzing the results of simulations in detail requires both tools to ask
sophisticated questions of data and the ability to process vast quantities of
data at high time-cadence.  As simulations grow in size and complexity, storing data
for post-processing simply becomes intractable.  To cope with this, we have
instrumented Enzo with the ability to conduct analysis during the course of a
simulation.  This enables analysis with extremely high time cadence (even as often
as every subcycle of the finest refinement level), as well as without
attempting to write an entire checkpoint output to disk.  The current mechanism
for conducting analysis in Enzo during the course of the simulation utilizes
the same compute resources as used by the simulation itself; this is often
referred to as \textit{in situ} analysis or visualization.  Utilizing a
dynamically-scheduled second set of computing resources, often referred to as
\textit{co-scheduled} analysis or visualization, provides greater flexibility and overall
throughput at the expense of simplicity.

@@

\subsubsection{Tracer particles}

@@

At user-specified intervals, the tracer particles output their data
in a streaming format into a binary file.  The data written out
typically includes the tracer particle's unique ID, position, velocity, density,
and temperature, but is easily extensible to output any grid-based
quantity that a user requires.  This capability has been used quite
effectively in several papers, including \citet{2010ApJ...715.1575S}
and \citet{2012ApJ...748...12S}. It is crucial to keep in mind that
tracer particles model some Lagrangian fluid trajectories. The fluid
on the grid, however, models the motion of all the mass and represents
the average quantity in a grid cells volume. Consequently, after some
evolution tracer particles, which even if they initially had the same
density distribution as the gas, will not have the same density
distribution as the fluid. They tend to accumulate at stagnation points
of the flow, e.g., and care has to be taken in using these particles
appropriately. They can be very useful, e.g. to study a variety of
histories of the hydrodynamic quantities in Lagrangian fluid elements
and evaluate complex chemical and cooling models in regard to the
simpler ones used in the actual numerical evolution.

\subsubsection{Shock finding}

>>>>

\section{Analysis}
\label{sec.num.analysis}

\subsection{Inline analysis with yt}

Analyzing the results of simulations in detail requires both tools to ask
sophisticated questions of data and the ability to process vast quantities of
data at high time-cadence.  As simulations grow in size and complexity, storing data
for post-processing simply becomes intractable.  To cope with this, we have
instrumented Enzo with the ability to conduct analysis during the course of a
simulation.  This enables analysis with extremely high time cadence (even as often
as every subcycle of the finest refinement level), without
attempting to write an entire checkpoint output to disk.  The current mechanism
for conducting analysis in Enzo during the course of the simulation utilizes
the same compute resources as used by the simulation itself; this is often
referred to as \textit{in situ} analysis or visualization.  Utilizing a
dynamically-scheduled second set of computing resources, often referred to as
\textit{co-scheduled} analysis or visualization, provides greater flexibility and overall
throughput at the expense of simplicity.

@@

\subsection{Tracer particles}

@@

At user-specified intervals, the tracer particles output their data.  The data written out
typically includes the tracer particle's unique ID, position, velocity, density,
and temperature, but is easily extensible to output any grid-based
quantity that a user requires.  This capability has been used quite
effectively in several papers, including \citet{2010ApJ...715.1575S}
and \citet{2012ApJ...748...12S}. It is crucial to keep in mind that
tracer particles model some Lagrangian fluid trajectories. The fluid
on the grid, however, models the motion of all the mass and represents
the average quantity in a grid cells volume. Consequently, after some
evolution tracer particles, which even if they initially had the same
density distribution as the gas, will not have the same density
distribution as the fluid. They tend to accumulate at stagnation points
of the flow, e.g., and care has to be taken in using these particles
appropriately. They can be very useful, e.g. to study a variety of
histories of the hydrodynamic quantities in Lagrangian fluid elements
and evaluate complex chemical and cooling models in regard to the
simpler ones used in the actual numerical evolution.

\subsection{Shock finding}

### 2013-05-19 numerical-conduction.tex
Added most of Dave Collins' suggested edits (except for a bunch of stuff having to do with the hydro).

where $\lambda_e$ is the electron mean free path and $\ell_T \equiv T
/ |\nabla T|$ is the characteristic length scale of the local
temperature gradient.  Furthermore, it is important to note that the
overall level of thermal conductivity can be strongly affected
by the presence of magnetic fields, which may strongly suppress heat
flow.  The strength and distribution of magnetic fields in the
intracluster medium is poorly understood, and it is unclear to what
extent these magnetic fields will suppress conduction.  For the
purposes of the work described in this paper, we assume that the
conductivity of the plasma can be described in terms of an effective
conductivity, which can be expressed as a fraction f$_{Sp}$ of the
Spitzer conductivity (where f$_{Sp} \leq 1.0$ are considered physical
values).

>>>>

where $\lambda_e$ is the electron mean free path and $\ell_T \equiv T
/ |\nabla T|$ is the characteristic length scale of the local
temperature gradient.  We also assume that the conductivity of the
plasma can be described in terms of an effective conductivity, which
can be expressed as a fraction f$_{Sp}$ of the Spitzer conductivity
(where f$_{Sp} \leq 1.0$ are considered physical values).  This takes
into account physical processes below the resolution limit of the
simulation (such as tangled magnetic fields) that can suppress heat
transport.

### 2013-05-19 numerical-gravity.tex
Added most of Dave Collins' suggested edits (except for a bunch of stuff having to do with the hydro).

First, the dark matter particles are distributed onto the grids using the second-order cloud-in-cell (CIC) interpolation technique \citep{Hockney88} to form a spatially-discretized density field $\rho_{\rm DM}$.  When doing the CIC interpolation, particle positions are (temporarily) advanced by $0.5 \Delta t v^n$ so that we generate an estimate of the time-centered density field.  Particles on sub-grids are also added in this way.  In addition, since the gravitating field for a grid is defined beyond the grid edges (see below), dark matter particles from sibling grids and sub grids whhatich lie within the entire gravitating field are used.  This step can involve communication.

>>>>

First, the dark matter particles are distributed onto the grids using the second-order cloud-in-cell (CIC) interpolation technique \citep{Hockney88} to form a spatially-discretized density field $\rho_{\rm DM}$.  When doing the CIC interpolation, particle positions are (temporarily) advanced by $0.5 \Delta t v^n$ so that we generate an estimate of the time-centered density field.  Particles on sub-grids are also added in this way.  In addition, since the gravitating field for a grid is defined beyond the grid edges (see below), dark matter particles from sibling grids and sub grids that lie within the entire gravitating field are used.  This step can involve communication.

### 2013-05-19 numerical-particle.tex
Added most of Dave Collins' suggested edits (except for a bunch of stuff having to do with the hydro).

\enzo\ uses a particle-mesh N-body method to calculate the dynamics of
collisionless systems \citep{Hockney88}.  This method follows
trajectories of a representative sample of individual particles that
sample the phase space of the dark matter distribution, and is much
more efficient than a direct solution of the Boltzmann equation in
essentially all astrophysical situations for the levels of accuracy
that are required for simulations of structure formation.  As
described earlier, the gravitational potential is computed by solving
the elliptic Poisson's equation (Eq.~\ref{eq:potential}), and
differencing the potential to find accelerations, which are then
interpolated back to particles.  This acceleration is time-centered
(because the underlying gravitating mass field is approximately
time-centered), and so we have accelerations $\myvec{a}^{n+1/2}$ for
each particle.  These are used to update the particle positions and
velocities starting from $\myvec{x}^n$ and $\myvec{v}^n$ using a
standard drift-kick-drift technique:

\begin{eqnarray}
\label{eqn.driftkick}
\myvec{x}^{n+1/2} & = & \myvec{x}^n + \frac{\Delta t}{2 a^n} \myvec{v}^{n} \nonumber \\
\myvec{v}^{n+1} & = & \myvec{v}^n \left(1 - \frac{\dot{a}^{n+1/2}}{a^{n+1/2}}\right) + \frac{\Delta t}{ a^{n+1/2}} \myvec{a}^{n+1/2} \\
\myvec{x}^{n+1} & = & \myvec{x}^{n+1/2} + \frac{\Delta t}{2 a^{n+1}} \myvec{v}^{n+1} \nonumber
\end{eqnarray}

>>>>

\enzo\ uses a particle-mesh N-body method to calculate the dynamics of
collisionless systems \citep{Hockney88}.  This method follows
trajectories of a representative sample of individual particles that
sample the phase space of the dark matter distribution, and is much
more efficient than a direct solution of the Boltzmann equation in
essentially all astrophysical situations for the levels of accuracy
that are required for simulations of structure formation.  As
described earlier, the gravitational potential is computed by solving
the elliptic Poisson's equation (Eq.~\ref{eq:potential}), and
differencing the potential to find accelerations, which are then
interpolated back to particles.  This acceleration is time-centered
(because the underlying gravitating mass field is approximately
time-centered), and so we have accelerations $\myvec{g}^{n+1/2}$ for
each particle.  These are used to update the particle positions and
velocities starting from $\myvec{x}^n$ and $\myvec{v}^n$ using a
standard drift-kick-drift technique:

\begin{eqnarray}
\label{eqn.driftkick}
\myvec{x}^{n+1/2} & = & \myvec{x}^n + \frac{\Delta t}{2 a^n} \myvec{v}^{n} \nonumber \\
\myvec{v}^{n+1} & = & \myvec{v}^n \left(1 - \frac{\dot{g}^{n+1/2}}{g^{n+1/2}}\right) + \frac{\Delta t}{ a^{n+1/2}} \myvec{g}^{n+1/2} \\
\myvec{x}^{n+1} & = & \myvec{x}^{n+1/2} + \frac{\Delta t}{2 a^{n+1}} \myvec{v}^{n+1} \nonumber
\end{eqnarray}
